TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-06-10 paper/draft.tex
discussion

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
%AWM: seems to me like the context here is all 3. You want to run on ~10,000 stars, which is expensive if you use models. You have labels for a small fraction of these. 
%The context of this paper is the last:
There are empirically-calibrated methods to measure \teff\ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of \teff\ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}. \thecannon\ fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
%AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models. The temperature tools use models, but they are calibrated against LBOI measurements. I need to come back to editing later, but this paragraph probably needs to be tweaked a bit. 
%AWM: OK, let me know what you think about those changes. Feel free to edit this more, but do note the work of Lindgren modeling J-band at high resolution (I think they use CRIRES).

@@

We trained and evaluated the performance of \thecannon\ at deriving atmospheric temperatures and metallicities from M dwarfs using high resolution near-infrared spectra, and compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures, metallicities, and spectral types, 6D kinematics (\gaia\ parallaxes and proper motions + \apogee\ radial velocities) shown in Table \ref{table:test_results}.

Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}, summarize which features are found to be strong temperature and metallicity indicators.

While \thecannon\ generally achieves excellent precision at predicting labels and reproducing the spectral features as demonstrated in this study and previous work on \thecannon\ \citep{Ness:2015,Casey:2016}, the accuracy of labels from model application to new sources is limited by the accuracy, relative precision, size, dynamic range and representation of the training sample.

@@

>>>>

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
%AWM: seems to me like the context here is all 3. You want to run on ~10,000 stars, which is expensive if you use models. You have labels for a small fraction of these. 
%The context of this paper is the last:
There are empirically-calibrated methods to measure $\teff$ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}. \thecannon\ fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
%AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models. The temperature tools use models, but they are calibrated against LBOI measurements. I need to come back to editing later, but this paragraph probably needs to be tweaked a bit. 
%AWM: OK, let me know what you think about those changes. Feel free to edit this more, but do note the work of Lindgren modeling J-band at high resolution (I think they use CRIRES).

@@

We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
from high-resolution infrared spectra from \apogee.
This work was motivated by the problem that M dwarfs stars are difficult to
model physically; the data are better than the models in important senses.
And indeed we find that our data-driven model is both accurate in the
data domain (as a spectral synthesis model) and precise in the latent domain
(as a tool for deriving physical parameters).
This accuracy and precision is consistent with previous work with
\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
extended to a new regime in spectral type ($\teff$) and surface
gravity.
The primary result of this work is that
we have compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures,
metallicities, spectral types, and six-dimensional kinematics.
These data are provided in Table \ref{table:test_results}.

While \thecannon\ achieves excellent precision at predicting labels
and reproducing spectral features, the
accuracy of labels it produces is limited by
the accuracy, relative precision, size, dynamic range and
representation of the training sample.
That is, being a supervised method, \thecannon\ is never any better in a mean (bias) sense
(although it can be better in a precision or variance sense) than the input training
data.
The catalog we have produced is a label transfer from parameters provided
in our input data (\citealt{Mann:2015}; M15) and it implicitly adopts all the biases and
issues from that input data.
It is also limited to the stellar-parameter domain of that input catalog.
That said, what we have done does also represent some kind of external validation of the M15
stellar parameters:
The success of our model shows that they do a great job of parameterizing a generative
model of the spectra.

@@

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in [BIRKY WHAT FIGURE] and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will, in the end, lead to
new and improved physical models, which will, in turn, put \thecannon\ out of business.
